Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
starting working on a top-down explanation of beta strands

HMMER simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
SMURFLite~\cite{smurflite}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Beta strands are regions of protein 
structure that ``stick'' to one another.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

>>>>

Our implementations of hidden Markov models and Viterbi's algorithm
simply duplicate 
functionality that is already implemented in the C++~package
HMMER.
These implementations form is just one component of our new tool, MRFy.

A hidden Markov model effectively uses local information about
amino-acid sequences.
But because real proteins are three-dimensional structures,
their amino-acid sequences are subject to \emph{nonlocal}
constraints. 
In~particular, when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can nevertheless be
adjacent in three-dimensional space.
Moreover, two such groups can become hydrogen-bonded to one another,
effectively ``stuck together.''
Such groups are called \emph{beta strands}.
\remark{I hope this is relatively accurate. --NR}

\emph{[Next up: how does this bonding phenomenon constrain sequences
of homologous proteins?]}

 simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
SMURFLite~\cite{smurflite}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Beta strands are regions of protein 
structure that ``stick'' to one another.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.
